State of the Practice for Medical Imaging Software

### 2022-10-31 StateOfPractice/Papers/MedImageSoft_SOP-arXiv/MedImageSoft_SOP.tex
Revisions to Pain points section

We present the state of the practice for Medical Imaging (MI) software. We
selected 29 medical imaging projects from 48 candidates, assessed 10 software
qualities (installability, correctness/ verifiability, reliability, robustness,
usability, maintainability, reusability, understandability,
visibility/transparency and reproducibility) by answering 108 questions for each
software project, and interviewed 8 of the 29 development teams. Based on the
quantitative data for the first 9 qualities, we ranked the MI software with the
Analytic Hierarchy Process (AHP). The four top ranked software products are:
\textit{3D Slicer}, \textit{ImageJ}, \textit{Fiji}, and \textit{OHIF Viewer}.
Our ranking is mostly consistent with the community's ranking, with four of our
top five projects also appearing in the top five of a list ordered by
stars-per-year. \textit{DICOM Viewer} and \textit{dmv} are exceptions; we ranked
them lower than their estimated popularity because of difficulty with their
installation.  Generally, MI software is in a healthy state.  The documentation
artifacts created, and development tools employed, typically match the
recommendations given in research software development guidelines. A few places
in the current state of the practice that deviate from the recommendations
include some underemphasized artifacts (like a developer's code of conduct, code
style guidelines, product roadmaps, and Application Program Interface (API)
documentation) and ... Exceptions include ... tools With respect to software
artifacts, MI software generally Comparison for tool usage. Comparison for
process. From interviewing the developers, we identified five pain points and
two qualities of potential concern:: i) lack of resources; ii) difficulty
balancing between compatibility, maintainability, performance, and security;
and, iii) lack of access to real-world datasets for testing. For future MI
software projects, we propose adopting test-driven development, using continuous
integration and continuous delivery (CI/CD), using git and GitHub, maintaining
good documentation, supporting third-party plugins or extensions, considering
web application solutions, and improving collaboration between different MI
software projects.
\wss{Update after the paper has been revised.}

@@

Our observations of an informally defined process, with elements of agile
methods, matches what has been observed for research software in general.
Scientific developers naturally use an agile philosophy \citep{AckroydEtAl2008,
CarverEtAl2007, EasterbrookAndJohns2009, Segal2005, HeatonAndCarver2015}, or an
amethododical process \citep{Kelly2013}, or a knowledge acquisition driven
process \citep{Kelly2015}.  A waterfall like process can work for research
software \citep{Smith2016}, especially if the developers work iteratively and
incrementally, but externally document their work as if they followed a
rationale design process \citep{parnas1986rational}.

@@

Our interviews identified pain points related to a lack of time and funding,
technology hurdles, improving correctness, and improving usability.  In this
section, we go through each pain point and contrast the MI experience with
observations from other domains.  We also cover potential ways to address the
pain points, as promoted by the community.  (Later, in
Section~\ref{ch_recommendations}, we propose additional pain mitigation
strategies based on our experience.)  In addition to pain points, we summarize
MI developer strategies for improving maintainability and reproducibility.
Although the interviewees did not explicitly identify these two qualities as
pain points, they were discussed as part of our interview process
\citep{SmithEtAl2021}.  The interviewee's practices for addressing pain points
and improving quality can potentially be emulated by other MI developers.
Moreover, these practices may provide examples that can be followed by other
research software domains.

@@

\begin{itemize}
\item Shifting from development to maintenance when the team does not have
enough developers for building new features and fixing bugs at the same time;
\item Improving documentation to save time answering users' and developers'
questions;
\item Supporting third-party plugins and extensions; and,
\item Using GitHub Actions for continuous integration and continuous
development.
\end{itemize}

\item[P\refstepcounter{pnum}\thepnum \label{P_LackFunding}:] \textbf{Lack of
Funding:} Developers felt the pain of having to attract funding to develop and
maintain their software. For instance, the interviewees from \textit{3D Slicer}
and \textit{OHIF} said getting funding for software maintenance is more
challenging than finding funding for research. The interviewee from the
\textit{ITK-SNAP} team thought more funding was a way to solve the lack of time
problem, because they could hire more dedicated developers. On the other hand,
the interviewee from the \textit{Weasis} team did not feel that funding could
solve the same problem, since they would still need time to supervise the project.

@@

The pain point survey of \citet{WieseEtAl2019} highlights that technology
hurdles are an issue for research software in general.  Some of the
technical-related problems mentioned by \citet{WieseEtAl2019} include dependency
management, cross-platform compatibility (also mentioned by
\citet{PintoEtAl2018}), CI, hardware issues and operating system issues. From
\citep{SmithEtAl2022} technology pain points for LBM developers include setting
up parallelization and CI.

@@

\begin{itemize}
\item Adopting a web-based approach with backend servers, to better support
lower-end computers;
\item Using memory-mapped files to consume less computer memory, to better
support lower-end computers; 
\item Using computing power from the computers GPU for web applications;
\item Maintaining better documentations to ease the development and maintenance
processes;
\item Improving performance via more powerful computers, which one interviewee
pointed out has already happened to reduce the balance problem.
\end{itemize}

@@

The advantage for native applications is higher performance, while web
applications have the advantage of cross-platform compatibility and a simpler
build process.  These web advantages mirror the native disadvantages of
difficulty with cross-platform compatibility and a complex build process.  The
lower performance disadvantage of web applications can be improved with a server
backend, but in this case there are disadvantages for privacy protection and
server costs.

@@

As implied by the above threats to correctness, testing was the most often
mentioned strategy for MI developers for ensuring correctness.  Seven teams
mentioned test related activities, including test-driven development, component
tests, integration tests, smoke tests, regression tests, self tests and
automated tests.  With the common emphasis on testing to improve correctness, MI
software is ahead of some other scientific domains.  For scientific software in
general \citet{PintoEtAl2018} mention the problem of insufficient testing and
\citet{HannayEtAl2009} show that more developers think testing is important than
the number that believe they have a sufficient understanding of testing
concepts.  Our study of LBM software suggests that this domain shares the
challenges of insufficient testing and insufficient understanding of testing
concepts \citep{SmithEtAl2022}. Automated testing is a specific challenge for
LBM software since free testing services do not offer adequate facilities for
large amounts of data \citep{SmithEtAl2022}. Although not specifically mentioned
during our interviews, the large data sets for MI likely cause a challenge for
using free testing services, like GitHub Actions.

Research software in general often struggles with the oracle problem for testing
because for many potential test cases the developer doesn't have a means to
judge the correctness of their calculated solutions \citep{HannayEtAl2009,
KanewalaAndBieman2013, KellyEtAl2011, WieseEtAl2019}.  The MI developers did not
allude to this challenge, likely because for a give image (test case) it is
possible to determine, possibly using other software, the expected analysis
results.

@@

Other suggested and practiced strategies include a graphical user interface,
testing every release with active users, making simple things simple and
complicated things possible, focusing on limited number of functions, icons with
clear visual expressions, designing the software to be intuitive, having a UX
designer, dialog windows for important notifications, providing an example for
users to follow, downsampling images to consume less memory, and an option to
load only part of the data to boost performance.  The last two points recognize
that an important component of usability is performance, since poor performance
frustrates users.

@@

\begin{enumerate}
\item[Q\refstepcounter{qnum}\theqnum \label{Q_Maintainability}:]
\textbf{Maintainability:} Although not explicitly highlighted as a pain point
during our interviews, \citet{Nguyen-HoanEtAl2010} rate maintainability as the
third most important software quality for research software in general. The push
for sustainable software \citep{deSouzaEtAl2019} is motivated by the pain that
past developers have had with accumulating too much technical debt
\citep{KruchtenEtAl2012}.  For LBM software, \citet{SmithEtAl2022} identifies
technical debt as one of the developer pain points.

@@

\item[Q\refstepcounter{qnum}\theqnum \label{Q_Reproducibility}:]
\textbf{Reproducibility:}  Although the MI developers did not mention
reproducibility explicitly as a pain point, they did mention the need to improve
documentation.  Good documentation does not just address the pain points of lack
of developer time (\ppref{P_LackDevTime}), technology hurdles
(\ppref{P_TechnologyHurdles}), usability \ppref{P_Usability}, and
Maintainability.  Documentation is also necessary for reproducibility. The
challenges of inadequate documentation are a known problem for research software
\citep{PintoEtAl2018, WieseEtAl2019} and for non-research software
\citep{LethbridgeEtAl2003}.

>>>>

We present the state of the practice for Medical Imaging (MI) software. We
selected 29 medical imaging projects from 48 candidates, assessed 10 software
qualities (installability, correctness/ verifiability, reliability, robustness,
usability, maintainability, reusability, understandability,
visibility/transparency and reproducibility) by answering 108 questions for each
software project, and interviewed 8 of the 29 development teams. Based on the
quantitative data for the first 9 qualities, we ranked the MI software with the
Analytic Hierarchy Process (AHP). The four top ranked software products are:
\textit{3D Slicer}, \textit{ImageJ}, \textit{Fiji}, and \textit{OHIF Viewer}.
Our ranking is mostly consistent with the community's ranking, with four of our
top five projects also appearing in the top five of a list ordered by
stars-per-year. \textit{DICOM Viewer} and \textit{dmv} are exceptions; we ranked
them lower than their estimated popularity because of difficulty with their
installation.  Generally, MI software is in a healthy state.  The documentation
artifacts created, and development tools employed, typically match the
recommendations given in research software development guidelines. A few places
in the current state of the practice that deviate from the recommendations
include some underemphasized artifacts (like a developer's code of conduct, code
style guidelines, product roadmaps, and Application Program Interface (API)
documentation) and ... Exceptions include ... tools With respect to software
artifacts, MI software generally Comparison for tool usage. Comparison for
process. From interviewing the developers, we identified five pain points and
two qualities of potential concern:: i) lack of resources; ii) difficulty
balancing between compatibility, maintainability, performance, and security;
and, iii) lack of access to real-world datasets for testing. For future MI
software projects, we propose adopting test-driven development, using Continuous
Integration and Continuous Delivery (CI/CD), using git and GitHub, maintaining
good documentation, supporting third-party plugins or extensions, considering
web application solutions, and improving collaboration between different MI
software projects.
\wss{Update after the paper has been revised.}

@@

Our observations of an informally defined process, with elements of agile
methods, matches what has been observed for research software in general.
Scientific developers naturally use an agile philosophy \citep{AckroydEtAl2008,
CarverEtAl2007, EasterbrookAndJohns2009, Segal2005, HeatonAndCarver2015}, or an
amethododical process \citep{Kelly2013}, or a knowledge acquisition driven
process \citep{Kelly2015}.  A waterfall-like process can work for research
software \citep{Smith2016}, especially if the developers work iteratively and
incrementally, but externally document their work as if they followed a
rationale design process \citep{parnas1986rational}.

@@

Our interviews identified pain points related to a lack of time and funding,
technology hurdles, improving correctness, and improving usability.  In this
section, we go through each pain point and contrast the MI experience with
observations from other domains.  We also cover potential ways to address the
pain points, as promoted by the community.  (Later, in
Section~\ref{ch_recommendations}, we propose additional pain mitigation
strategies based on our experience.)  In addition to pain points, we summarize
MI developer strategies for improving maintainability and reproducibility.
Although the interviewees did not explicitly identify these two qualities as
pain points, we did discuss threats to these qualities and ways to improve them
as part of our interview process \citep{SmithEtAl2021}.  The interviewee's
practices for addressing pain points and improving quality can potentially be
emulated by other MI developers. Moreover, these practices may provide examples
that can be followed by other research software domains.

@@

\begin{itemize}
\item Shifting from development to maintenance when the team does not have
enough developers for building new features and fixing bugs at the same time;
\item Improving documentation to save time answering users' and developers'
questions;
\item Supporting third-party plugins and extensions; and,
\item Using GitHub Actions for CI/CD (Continuous Integration and Continuous
Delivery.)
\end{itemize}

\item[P\refstepcounter{pnum}\thepnum \label{P_LackFunding}:] \textbf{Lack of
Funding:} Developers felt the pain of having to attract funding to develop and
maintain their software. For instance, the interviewees from \textit{3D Slicer}
and \textit{OHIF} said getting funding for software maintenance is more
challenging than finding funding for research. The interviewee from 
\textit{ITK-SNAP} thought more funding was a way to solve the lack of time
problem, because they could hire more dedicated developers. On the other hand,
the interviewee from \textit{Weasis} did not feel that funding could
solve the same problem, since they would still need time to supervise the project.

@@

The pain point survey of \citet{WieseEtAl2019} highlights that technology
hurdles are an issue for research software in general.  Some technical-related
problems mentioned by \citet{WieseEtAl2019} include dependency management,
cross-platform compatibility (also mentioned by \citet{PintoEtAl2018}), CI,
hardware issues and operating system issues. From \citep{SmithEtAl2022}
technology pain points for LBM developers include setting up parallelization and
CI.

@@

\begin{itemize}
\item Adopting a web-based approach with backend servers, to better support
lower-end computers;
\item Using memory-mapped files to consume less computer memory, to better
support lower-end computers; 
\item Using computing power from the computers GPU for web applications;
\item Maintaining better documentations to ease the development and maintenance
processes;
\item Improving performance via more powerful computers, which one interviewee
pointed out has already happened.
\end{itemize}

@@

The advantage for native applications is higher performance, while web
applications have the advantage of cross-platform compatibility and a simpler
build process.  These web advantages mirror the native disadvantages of
difficulty with cross-platform compatibility and a complex build process.  The
lower performance disadvantage of web applications can be improved with a server
backend, but in this case there are disadvantages for privacy protection and
server costs.  These issues are discussed further in the recommendations
(Section~\ref{sec_webapps}).

@@

As implied by the above threats to correctness, testing was the most often
mentioned strategy for MI developers for ensuring correctness.  Seven teams
mentioned test related activities, including test-driven development, component
tests, integration tests, smoke tests, regression tests, self tests and
automated tests.  With the common emphasis on testing to improve correctness, MI
software is ahead of some other scientific domains.  For scientific software in
general \citet{PintoEtAl2018} mention the problem of insufficient testing and
\citet{HannayEtAl2009} show that more developers think testing is important than
the number that believe they have a sufficient understanding of testing
concepts.  Our study of LBM software suggests that this domain shares the
challenges of insufficient testing and insufficient understanding of testing
concepts \citep{SmithEtAl2022}. Automated testing is a specific challenge for
LBM software since free testing services do not offer adequate facilities for
large amounts of data \citep{SmithEtAl2022}. Although not specifically mentioned
during our interviews, the large data sets for MI likely also cause a challenge
for using free testing services, like GitHub Actions.

Research software in general often struggles with the oracle problem for testing
because for many potential test cases the developer doesn't have a means to
judge the correctness of their calculated solutions \citep{HannayEtAl2009,
KanewalaAndBieman2013, KellyEtAl2011, WieseEtAl2019}.  The MI developers did not
allude to this challenge, likely because for a give image (test case) it is
possible to determine, potentially by using other software, the expected
analysis results.

@@

Other suggested and practiced strategies include a graphical user interface,
testing every release with active users, making simple things simple and
complicated things possible, focusing on limited number of functions, icons with
clear visual expressions, designing the software to be intuitive, having a UX
(User eXperience) designer, dialog windows for important notifications,
providing an example for users to follow, downsampling images to consume less
memory, and providing an option to load only part of the data to boost
performance.  The last two points recognize that an important component of
usability is performance, since poor performance frustrates users.

@@

\begin{enumerate}
\item[Q\refstepcounter{qnum}\theqnum \label{Q_Maintainability}:]
\textbf{Maintainability:} \citet{Nguyen-HoanEtAl2010} rate maintainability as the
third most important software quality for research software in general. The push
for sustainable software \citep{deSouzaEtAl2019} is motivated by the pain that
past developers have had with accumulating too much technical debt
\citep{KruchtenEtAl2012}.  For LBM software, \citet{SmithEtAl2022} identifies
technical debt as one of the developer pain points.

@@

\item[Q\refstepcounter{qnum}\theqnum \label{Q_Reproducibility}:]
\textbf{Reproducibility:}  Although the MI developers did not mention
reproducibility explicitly as a pain point, they did mention the need to improve
documentation.  Good documentation does not just address the pain points of lack
of developer time (\ppref{P_LackDevTime}), technology hurdles
(\ppref{P_TechnologyHurdles}), usability \ppref{P_Usability}, and
maintainability.  Documentation is also necessary for reproducibility. The
challenges of inadequate documentation are a known problem for research software
\citep{PintoEtAl2018, WieseEtAl2019} and for non-research software
\citep{LethbridgeEtAl2003}.
